---
title: "Multivariate Data Analysis: Airline Data"
author: "Huong Ngo"
date: "3/17/2022"
output: beamer_presentation
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(warning = FALSE)

# Loading libraries
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(ggrepel)
library(GGally)
set.seed(302)

# Importing datasets
profit_df <- read_csv("/Users/huongngo/Desktop/SPA DRP - Research/eda-airline-data/Data/profit_data.csv")
traffic_capacity_df <- read_csv("/Users/huongngo/Desktop/SPA DRP - Research/eda-airline-data/Data/traffic_capacity_data.csv")

# Graph theme
g_theme <- theme_bw(base_size = 16) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"),
        plot.subtitle = element_text(hjust = 0.5))
```

## Outline

- What is Multivariate Data Analysis?
- The Dataset
- Preliminary Data Visualizations
- Principal Components Analysis (PCA)
- K-Means Clustering
- Introducing Other Methods for Future Work
- What I Learned!

## What is Multivariate Data Analysis

- From *STAT 311*, students learned about **univariate data analysis**: analysis on data that is of **one (univariate)** variable
- Multivariate data: data that **contains values recorded for** *multiple different* **variables**
- Multivariate data analysis: *simultaneous* **statistical analysis of a collection of variables**
- Main goal: **uncover, display or extract any "signal" in the data in the presence of noise to discover what the data has to tell us**

## The Dataset

- Examined operations financial data of passenger airline carriers from 2019 - 2021
- 26 variables
  - Revenue sources: *Baggage Fees, Cancellation Fees, Transportation Services*
  - Expense sources: *Maintenance, Aircraft Services, Passenger Services*
  - Overall information: *Net Income, Operation Profit/Loss, Operation Revenues/Expenses*
- 496 observations
  - Each observation is the financial details of an airline carrier in a quarter of a specific year in a specific region
- Excluded cargo and charter airliners

## Preliminary Data Visualizations

```{r}
# Scatterplot Matrix for profit_df
profit_scatter_df <- profit_df %>% select(prop_bag,
                                          trans_rev_pax,
                                          op_revenues,
                                          flying_ops,
                                          maintenance,
                                          op_expenses)
plot(profit_scatter_df)
```

## Motivations from Data Visualizations and Nature of Dataset

- Too many variables!
- Correlation between variables in dataset: Collinearity!
- Solution:
  - **PCA**: Reduce dimensionality of dataset while retaining as much variation of the original dataset
- Clusters in Scatterplot Matrix
  - **K-Means Clustering**: Cluster analysis that aims to uncover clusters of observations that are homogeneous

## Principal Components Analysis (PCA)

- *princomp()*: PCA function that performs a series of calculations to describe variation of a set of correlated variables as a new, uncorrelated set of variables.
  - New set of variables are the linear combinations of the old variables
  - We will use the first few variables (2 or 3) as they will capture the most variation and give a lower-dimensional summary of data. These are called the principal components
- Standardizing dataset: The variables have different range of values
  - To ensure our analysis is not affected by the scales, we will standardize it
  
---

## Principal Components Analysis (PCA)

```{r}
# Retrieving dataframe with only numerical values
profit_pca_df <- profit_df[-(19:26)]
# Running PCA
profit_pca_list <- princomp(profit_pca_df, cor = TRUE)

# Plotting the loadings in two dimensional space
profit_pca_list$loadings <- unclass(profit_pca_list$loadings)

profit_loadings <- as.data.frame(profit_pca_list$loadings)

# Plotting the principal components in two dimensional space
profit_scores <- as.data.frame(profit_pca_list$scores)
profit_scores <- profit_scores %>% 
  mutate(carrier_name = profit_df$unique_carrier_name,
         year = as.factor(profit_df$year),
         quarter = as.factor(profit_df$quarter),
         region = as.factor(profit_df$region))
```


```{r}
# Generating scree plot to choose number of components
screeplot(profit_pca_list, type = "lines")
```

## Interpretation of Principal Components Analysis (PCA)
- First Principal Component is characterized by **revenue and expense sources**
- Second Principal Component is characterized by **general financial information** about a carrier
- Third Principal Component is characterized by **revenue gained from moving freight**

## Principal Components Analysis (PCA)

```{r}
# PC1 and PC2 - Colored by quarter, Faceted by year
# 2019
ggplot(data = profit_scores %>% filter(year == 2019), aes(x = Comp.1,
                                                          y = Comp.2,
                                                          label = carrier_name,
                                                          color = quarter)) +
  geom_point(size = 3,
             alpha = 0.5) +
  geom_text_repel(size = 4,
                  color = "red",
                  nudge_x = 0.02,
                  nudge_y = -0.02,
                  max.overlaps = getOption("ggrepel.max.overlaps", default = 5)) +
  geom_vline(xintercept = 0, color = "gray") +
  geom_hline(yintercept = 0, color = "gray") +
  labs(title = "First Two Principal Components",
       x = "First Principal Component",
       y = "Second Principal Component",
       color = "Quarter",
       subtitle = "Colored by quarter in year 2019") +
  g_theme
```

---

## Principal Components Analysis (PCA)

```{r}
# PC1 and PC2 - Colored by quarter, Faceted by year
# 2020
ggplot(data = profit_scores %>% filter(year == 2020), aes(x = Comp.1,
                                                          y = Comp.2,
                                                          label = carrier_name,
                                                          color = quarter)) +
  geom_point(size = 3,
             alpha = 0.5) +
  geom_text_repel(size = 4,
                  color = "red",
                  nudge_x = 0.02,
                  nudge_y = -0.02) +
  geom_vline(xintercept = 0, color = "gray") +
  geom_hline(yintercept = 0, color = "gray") +
  labs(title = "First Two Principal Components",
       x = "First Principal Component",
       y = "Second Principal Component",
       color = "Quarter",
       subtitle = "Colored by quarter in year 2020") +
  g_theme
```

---

## Principal Components Analysis (PCA)

```{r}
# PC1 and PC2 - Colored by quarter, Faceted by year
# 2021
ggplot(data = profit_scores %>% filter(year == 2021), aes(x = Comp.1,
                                                          y = Comp.2,
                                                          label = carrier_name,
                                                          color = quarter)) +
  geom_point(size = 3,
             alpha = 0.5) +
  geom_text_repel(size = 4,
                  color = "red",
                  nudge_x = 0.02,
                  nudge_y = -0.02,
                  max.overlaps = getOption("ggrepel.max.overlaps", default = 5)) +
  geom_vline(xintercept = 0, color = "gray") +
  geom_hline(yintercept = 0, color = "gray") +
  labs(title = "First Two Principal Components",
       x = "First Principal Component",
       y = "Second Principal Component",
       color = "Quarter",
       subtitle = "Colored by quarter in year 2021") +
  g_theme
```

## K-Means Clustering
- **kmeans()** - K-Means Clustering algorithms function that partitions n observations in a dataset into k clusters that minimizes a numerical criterion
  - Within-Group Sum of Squares (WGSS)
  - Intuitively, it is a value to measure the difference between the observations within a group
- Standardizing dataset, just like PCA!

---

## K-Means Clustering

```{r}
profit_scaled_df <- profit_df[, 1:18]
mu <- colMeans(profit_scaled_df)
sigma <- sapply(profit_scaled_df, sd)
for (i in 1:ncol(profit_scaled_df)) {
  for (j in 1:nrow(profit_scaled_df)) {
    profit_scaled_df[j, i] <- (profit_scaled_df[j, i] - mu[i])/sigma[i]
  }
}

# 2019
profit_scaled_df <- profit_scaled_df %>%
  mutate(year = profit_df$year)

profit_scaled_2019 <- profit_scaled_df %>% 
  filter(year == 2019)

wss_2019 <- rep(0, 6)
for (i in 2:6) {
  wss_2019[i] <- sum(kmeans(profit_scaled_2019, centers = i)$withinss)
}

# Getting all tested k values
wss_2019_df <- data.frame(wss_2019)
wss_2019_df <- wss_2019_df %>% 
  mutate(num_cluster = 1:6) %>%
  filter(rownames(wss_2019_df) != 1)

# Running k-means clustering on 2019 data
k_clusters_5_2019 <- kmeans(profit_scaled_2019, centers = 5)

# 2020
profit_scaled_2020 <- profit_scaled_df %>% 
  filter(year == 2020)

wss_2020 <- rep(0, 6)
for (i in 2:6) {
  wss_2020[i] <- sum(kmeans(profit_scaled_2020, centers = i)$withinss)
}

# Getting all tested k values
wss_2020_df <- data.frame(wss_2020)
wss_2020_df <- wss_2020_df %>% 
  mutate(num_cluster = 1:6) %>%
  filter(rownames(wss_2020_df) != 1)

# Running k-means clustering on 2020 data
k_clusters_4_2020 <- kmeans(profit_scaled_2020, centers = 4)

# 2021
profit_scaled_2021 <- profit_scaled_df %>% 
  filter(year == 2021)

wss_2021 <- rep(0, 6)
for (i in 2:6) {
  wss_2021[i] <- sum(kmeans(profit_scaled_2021, centers = i)$withinss)
}

# Getting all tested k values
wss_2021_df <- data.frame(wss_2021)
wss_2021_df <- wss_2021_df %>% 
  mutate(num_cluster = 1:6) %>%
  filter(rownames(wss_2021_df) != 1)

# Running k-means clustering on 2021 data
k_clusters_4_2021 <- kmeans(profit_scaled_2021, centers = 4)
```

```{r}
# Plotting PC1 and PC3
ggplot(data = profit_scores %>% filter(year == 2020), aes(x = Comp.1,
                                                          y = Comp.3,
                                                          label = carrier_name,
                                                          color = as.factor(k_clusters_4_2020$cluster))) +
  geom_point(size = 3,
             alpha = 0.5) +
  geom_text_repel(size = 4,
                  color = "red",
                  nudge_x = 0.02,
                  nudge_y = -0.02,
                  max.overlaps = getOption("ggrepel.max.overlaps", default = 5)) +
  geom_vline(xintercept = 0, color = "gray") +
  geom_hline(yintercept = 0, color = "gray") +
  labs(title = "First and Third Principal Components",
       x = "First Principal Component",
       y = "Third Principal Component") +
  theme_bw()
```

## Introducing Other Methods for Future Work
- Multidimensional Scaling
  - Similar to PCA in its aim!
  - Applied to **distance matrices** 
- Analysis of Repeated Measures
  - Applied to multivariate data that contains repeated measurements on the same variable on each unit
  - Aims to examine change in the repeated values of the response variable and determine any explanatory variables associated with it
  - An upgrade from usual linear regression to account for the "unobserved" effects

## What I Learned/Gained!
- Many new skills!
- Project-learning experience!
- Confidence to learn on my own and craft my own projects!
- A great mentor :)


